---
title: 'Assignment 2: Data visualization'
author: "Pesthy Zsuzsi"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r}

library(tidyverse)
library(tidytuesdayR)
library(readr)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}

#loading data
expeditions <- read_csv("C:/Users/Zsuzsi/BML-MEMO LAB Dropbox/Lab_workspace/PhD_msc_students/Pesthy_Zsuzsi/phd/nt stat/expeditions.csv")

```

```{r}
#top15, ordered 
top15_df <- expeditions %>%
  mutate(peak_name = forcats::fct_lump(peak_name, n = 15, other_level = "Other")) %>%
  filter(peak_name != "Other") %>% 
  group_by(peak_name) %>%
  mutate(total_members = sum(members)) %>%
  ungroup() %>%
  mutate(peak_name = forcats::fct_reorder(peak_name, total_members))

#lets see the plot
ggplot(top15_df, aes(x = peak_name, y = members, fill = season)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "The 15 most popular peaks stacked by season of expedition",
    x = "",
    y = "Number of expeditions",
    fill = "season") +
  theme_light() +
  scale_fill_viridis_d()


```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
#loading data
phd_data <- read_csv("C:/Users/Zsuzsi/BML-MEMO LAB Dropbox/Lab_workspace/PhD_msc_students/Pesthy_Zsuzsi/phd/nt stat/phd_by_field.csv", show_col_types = FALSE)

```

```{r}
#clean and aggregate data 
plot_df <- phd_data %>%
  filter(!is.na(year), !is.na(n_phds), !is.na(broad_field)) %>%
  group_by(year, broad_field) %>%
  summarise(n_phds = sum(n_phds), .groups = "drop")

#plot
ggplot(plot_df, aes(x = year, y = n_phds, color = broad_field)) +
  geom_line(linewidth = 1.2) +
  theme_light() +
  labs(title = "Number of awarded Ph.D.-s in the US by year",
    x = "",
    y = "",
    color = "Broad field") +
  #the colors are not ordered automatically, so I set them manually :(
  scale_color_manual(values = c(
    "Engineering" = "#D95F02",   
    "Life sciences" = "#E62588", 
    "Humanities and arts" = "#6E69AF", 
    "Mathematics and computer sciences" = "#60A215", 
    "Psychology and social sciences" = "#A6751C", 
    "Education" = "#109971",     
    "Other" = "#E6AB02")) +
  theme(
    panel.border = element_blank(),
    panel.grid = element_blank())+
  scale_x_continuous(breaks = seq(2008, 2016, 2),
                     limits = c(2008, 2016)) +
  scale_y_continuous(labels = scales::label_comma(big.mark = ","))


# I noticed the suggested solution only after finishing the plots...
# hopefully it's okay that I used a different approach
# (this is also how it's done on DataCamp)

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
#loading data


commute_data <- read_csv("C:/Users/Zsuzsi/BML-MEMO LAB Dropbox/Lab_workspace/PhD_msc_students/Pesthy_Zsuzsi/phd/nt stat/commute.csv", show_col_types = FALSE)


```

```{r}
# aggregate commute counts by state and mode
commute_data_wide <- commute_data %>%
  group_by(state_abb, mode) %>%
  mutate(total_commutes = sum(n, na.rm = T)) %>% 
  select(state_abb, state_region, mode, total_commutes) %>% 
  distinct() %>% 
  pivot_wider(names_from = mode, values_from = total_commutes)

#plot
ggplot(commute_data_wide, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  geom_text(aes(label = state_abb), size = 3, vjust = -1, color = "black", show.legend = FALSE)+
  theme_light() +
  labs(
    title = "Number of people walking vs. biking to work in each USA state",
    x = "Number of ppl biking to work (log N)",
    y = "Number of ppl walking to work (log N)",
    color = "State region") +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format(), limits = c(25, NA))

```
